---
title: "modeldb"
output:
  github_document:
    toc: true
    toc_depth: 3
---


```{r setup, include=FALSE}
library(dplyr)
library(modeldb)
```

Fit models inside the database. **`modeldb` works with several databases back-ends** because it leverages `dplyr` and `dbplyr` for the final SQL translation of the algorithm. It currently supports:

- K-means clustering

- Linear regression

## Installation

Install the development version using `devtools` as follows:
```{r, eval = FALSE}
devtools::install_github("edgararuiz/modeldb")
```

## Linear regression

An easy way to try out the package is by creating a temporary SQLite database, and loading `mtcars` to it

```{r}
con <- DBI::dbConnect(RSQLite::SQLite(), path = ":memory:")
RSQLite::initExtension(con)
dplyr::copy_to(con, mtcars)
```

```{r}
library(dplyr)

tbl(con, "mtcars") %>%
  select(wt, mpg, qsec) %>%
  linear_regression(wt)
```

The model output can be parsed by `tidypredict` to run the predictions in the database. Please see the `Linear Regression` article to learn more about how to use `linear_regression()`

## K Means clustering

To use the `simple_kmeans()` function, simply pipe the database back end table to the function.  This returns a list object that contains two items:

- A sql query table with the final center assignment
- A local table with the information about the centers

```{r}
km <- tbl(con, "mtcars") %>%
  simple_kmeans(mpg, wt)

km$centers
```

To preview the record level center assignment use the `tbl` item

```{r}
head(km$tbl)
```

The SQL statement from `tbl` can be extracted using `dbplyr`'s `remote_query()`

```{r}
dbplyr::remote_query(km$tbl)
```

More information can be found in the `KMeans Clustering` article.

## Supported models

The following R models are currently supported. For more info please review the corresponding vignette:

- [Linear Regression](http://modeldb.netlify.com/articles/linear_regression/) - `linear_regression()`
- [K-means clustering](http://modeldb.netlify.com/articles/kmeans/) - `simple_kmenas()`


